---
title: "Code Tree Document"
author: "Simon Bond"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_vignette:
    keep_md: no
classoption: landscape
header-includes:
- \usepackage{tikz}
- \usetikzlibrary{trees,arrows}
vignette: |
  %\VignetteIndexEntry{Code Tree Document} 
  %\VignetteEncoding{UTF-8} 
  %\VignetteEngine{knitr::rmarkdown}
---





```{r include=FALSE}
library(magrittr)
library(knitr)
PATH <- getwd()
#PATH <- normalizePath(paste0(PATH,"/.."))
print(getwd())

code_tree <-read.csv("Output/codetree.csv")
print(code_tree)


code_tree %<>% within({
  parent %<>% gsub("\\\\","/",., fixed=TRUE) %>% gsub(PATH,"",  . ,fixed=TRUE)
  child %<>% gsub("\\\\","/",., fixed=TRUE) %>% gsub(PATH,"",  . ,fixed=TRUE)
})

graphcode <- function( edges, node, rootnode=TRUE){
  #edges is a data.frame with two columns: parent, child
  if(length(node)){
    children <-  subset(edges, parent==node)$child
    if(!rootnode){ output <- " child{ "}
    if(rootnode){ output <- "" }
    output <- c(output,  "node{ ", node,"} ")
    #if(rootnode){ output <- c(output, "[grow=right] ")}
    for( child in children){ 
      output <- c(output, Recall(edges, child, rootnode=FALSE))
    }
    if(!rootnode){ output <- c( output, " }")}
  } else{ 
    output <- ""
  }
  output
}



graph <- graphcode(code_tree, code_tree[1,1])
graph <- gsub("\\\\","/", graph)
graph <- gsub(PATH,"", graph)
graph <- gsub("_"," ", graph)
graph <- paste(c('\\',graph, ";"), collapse = "")
graph

#sibling distance=20em,
```

\begin{tikzpicture}[edge from parent fork right, level distance=20em, grow'=right,
  every node/.style = {shape=rectangle, rounded corners,draw, align=center, fill=white}] 
     `r knit_child(text=graph)`
\end{tikzpicture}

# Real Usage

This is generated a stand alone vignette, but really it would be used in a modified form with the following YAML in the .Rmd file


```
---
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
header-includes:
  - \usepackage{tikz}
  - \usetikzlibrary{trees,arrows}
output: pdf_document
classoption: landscape
params:
  my_title: "You only live twice"
  my_author: "Ian Flemming"
title: "`` `r
params$my_title` ``"  
author: "`` `r
params$my_author` ``"
---
```

and rendered from the `main.R` with something like

```{r, eval=FALSE}
render("Progs/rmarkdown_report.Rmd", 
       output_file ="Vignette Report.html",output_dir = "Output/Reports",
       params=list(my_author="Simon Bond", 
                   my_title= "Vignette Report"),
       knit_root_dir=".."
)
```
