---
title: "Code Tree Document"
author: "Simon Bond"
date: "`r Sys.Date()`"
output:
  pdf_document: 
    latex_engine: xelatex
  html_vignette:
    keep_md: no
classoption: landscape
header-includes:
  - \usepackage{tikz}
  - \usetikzlibrary{trees,arrows}
vignette: |
  %\VignetteIndexEntry{Code Tree Document} 
  %\VignetteEncoding{UTF-8} 
  %\VignetteEngine{knitr::rmarkdown}
---





```{r include=FALSE}
library(magrittr)
library(knitr)

code_tree <- read.csv(file.path("Output","codetree.csv"), stringsAsFactors = FALSE)


root <- code_tree[1,1]

PATH <- gsub("\\\\ROOT$", "", root,perl=TRUE) 

code_tree %<>% within({
  parent %<>% gsub("\\\\","/",., fixed=TRUE) %>% gsub(PATH,"",  . ,fixed=TRUE)
  child %<>% gsub("\\\\","/",., fixed=TRUE) %>% gsub(PATH,"",  . ,fixed=TRUE)
})

graphcode <- function( edges, node, rootnode=TRUE){
  #edges is a data.frame with two columns: parent, child
  if(length(node)){
    children <-  subset(edges, parent==node)$child
    if(!rootnode){ output <- " child{ "}
    if(rootnode){ output <- "" }
    output <- c(output,  "node{ ", node,"} ")
    #if(rootnode){ output <- c(output, "[grow=right] ")}
    for( child in children){ 
      output <- c(output, Recall(edges, child, rootnode=FALSE))
    }
    if(!rootnode){ output <- c( output, " }")}
  } else{ 
    output <- ""
  }
  output
}



graph <- graphcode(code_tree, code_tree[1,1])
graph <- gsub("\\\\","/", graph)
graph <- gsub(PATH,"", graph)
graph <- gsub("_"," ", graph)
graph <- paste(c('\\',graph, ";"), collapse = "")
graph

#sibling distance=20em,
```

\begin{tikzpicture}[edge from parent fork right, level distance=20em, grow'=right,
  every node/.style = {shape=rectangle, rounded corners,draw, align=center, fill=white}] 
     `r knit_child(text=graph)`
\end{tikzpicture}

# Real Usage

This is generated a stand alone vignette, but really it woudl be called from a call in `main.R` like


```{r, eval=FALSE}
render("Progs/code_tree_doc.Rmd", 
       output_file ="Vignette Code Tree.pdf",output_dir = "Output/Reports",
       params=list(my_author="Simon Bond", 
                   my_title= "Vignette Code Tree")
       
)
```

NB use the file `/doc/Progs/code_tree_docs.Rmd` and not the similarly named one within `/doc` . Building vignettes requires these two slightly different versions. 

