---
title: "Use cttab in the Analysis"
author: "Alimu Dayimu"
date: "`r Sys.Date()`"
output:
  knitr:::html_vignette:
    toc: true
  pdf_document: 
header-includes:
- \usepackage{tikz}
- \usetikzlibrary{trees,arrows}
vignette: |
  %\VignetteIndexEntry{Use cttab in the Analysis} 
  %\VignetteEncoding{UTF-8} 
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(cctu)
```
# What this Document Does

The package `cctu` has been updated to utilize the DLU file and CLU file to facilitate the analyzing process. The newly added functions will not affect previous process. But here, in this document we are going to learn some new functions provided in the package to facilitate your analysis. Before you start, you should get yourself familiarized with the `cctu` package by looking through the `Analysis Template` vignette. **Remember, the `sumby` function will not benefit from any of the data attribute related new functions mentioned in this document**. Namely, the variable label and value labels will not have an effect on the output. You don't need to go through everything in this document if you are going to stick to the `sumby` function.

# What's new

## Variable and value label
If you are familiar with `SAS`, `Stata` or `SPSS`, you should already know there's a variable label and value label (variable format in some). The variable label gives you the description of the variable, and the value label is to explain what the values stand for. The variable will stay as a numeric but has categories attached to it. Which means, you can subset or manipulate it like a numerical variable but report the data as a categorical one. It is important to understand this, let's use `mtcars` dataset and demonstrate how it works.

```{r echo=TRUE, message=FALSE, warning=FALSE}
data(mtcars)
# Assign variable label
var_lab(mtcars$am) <- "Transmission"

# Assign value label with named vector
val_lab(mtcars$am) <- c("Automatic" = 0, "Manual" = 1)
str(mtcars$am)
```

As you can above, `label` and `labels` attributes were added. But the variable is numeric. You can still summarize it as numeric variable as below:
```{r echo=TRUE}
summary(mtcars$am)
```

You can convert the value label to factor just before your final analysis. The `to_factor` function replace the value with labels and convert the variable to factor.
```{r echo=TRUE}
# Extract variable label
var_lab(mtcars$am)

# Convert variable to factor with labels attached to it
table(to_factor(mtcars$am))
```

But be cautious, **some R process might drop the variable or value label attributes**. Normally it won't, but you should check it before report if you are not sure. For example, `as.numeric`, `as.character` and `as.logical` will drop the variable and value label. There are `to_numeric`, `to_character` and `to_logical` can be used to convert the data type. You can also use the `copy_lab` function to copy the variable and value labels from the other variable.

You can use `var_lab` to extract the variable label or assign one. And use `has_label` to check if the variable has any variable label with it. You may also want to use `drop_lab` to drop the variable label.    

For value label, you can use `val_lab` to extract or assign value label. And use `has_labels` to check if the variable has a value label. There are also `unval` function to drop the value label and `lab2val` to replace the `data.frame` value to its corresponding value labels.


## MACRO dataset utility functions

There are new functions have been added to utilize the DLU and CLU files for the data analysis. The `apply_macro_dict` function uses DLU and CLU file to assign variable and value labels to the dataset. This function will convert the variable name in the data and DLU to lower case by default. This will also convert the dataset based on the variable type as in the DLU file. If you don't want to convert the variable name to lower cases, you should set `clean_names = FALSE`. 

The `extract_form` can extract MACRO data by form and visits. The data will be converted to `data.table` class, which is an extension of the `data.frame` and works exactly the same. It is a great package with lots of data manipulation capability, you should seek the [website](https://CRAN.R-project.org/package=data.table) for more details. But one thing to remember is that all the names in the variable selection will be considered as a variable of the data. 

```{r echo=TRUE, eval=FALSE}
vars <- c("mpg", "am")
# You can do this in the normal data.frame
mtcars[, vars]

# But you can't do this for the data.table
dat <- data.table::data.table(mtcars)
dat[, vars]

# You need to add with=FALSE to do that
dat[, vars, with = FALSE]
```

## Table function

You might already knew how to use the `sumby` function, but now a new function called `cttab` has been added. The difference between these two functions are the latter can handle variable label and value labels. You can feed the labelled data to this function and it will populate a summary table. Report data by treatment group, stratify tables by visit. Also, you can report variable based on some conditions and group the variable in the report. It generates a missing report internally and you can dump the missing report at the end. No extra step is needed. It will also produce the summary plots of the variables. The produced plots will be arranged to 3 by 3 and a new plot will be produced if the variables exceed 9. The remaining of this document will show you with a working case.

## Setting options

The `cttab` function has a good flexibility, which means it has lots of parameters you can use. You should check out the manual of the `cttab` function. The recommended way to set defaults is `cctu_options()` (and `cctu_opt()` to read a single value back); these honour the same names you'd otherwise pass as `options(cctu_<name> = …)`, but with input validation:

```{r eval = FALSE}
cctu_options(
  digits        = 3,        # keep 3 significant digits for numerical value
  digits_pct    = 0,        # keep 0 digits for percentage
  subjid_string = "subjid", # subject ID column name
  print_plot    = FALSE     # don't produce summary plots
)

# Read a single option back
cctu_opt("digits")
```

The plain `options(cctu_<name> = …)` form is still supported for backwards compatibility.


# Working example

In this section, we will show how to populate tables.

## Data reading
### Usual case
You should read the data as before, but you can and should read the data by setting all the columns to character. This will be handled later. Same process can be found in the `Analysis Template` vignette.
```{r echo=TRUE}
# Read example data
dt <- read.csv(system.file("extdata", "pilotdata.csv", package = "cctu"), colClasses = "character")

# Read DLU and CLU
dlu <- read.csv(system.file("extdata", "pilotdata_dlu.csv", package = "cctu"))
clu <- read.csv(system.file("extdata", "pilotdata_clu.csv", package = "cctu"))
```


### Vertically split multiple data
If you have multiple datasets, you should combine them at this stage. If multiple datasets are given and they have the same variable name, you can use `rbind` function to combine data together. If you have different datasets with different variable names except for some key variables, you can use `merge_data` function to combine them together.

```{r eval=FALSE, echo=TRUE}
# Read example data
dt_a <- read.csv(system.file("extdata", "test_A.csv", package = "cctu"),
  colClasses = "character"
)
dt_b <- read.csv(system.file("extdata", "test_B.csv", package = "cctu"),
  colClasses = "character"
)

# Read DLU and CLU
dlu_a <- read.csv(system.file("extdata", "test_A_DLU.csv", package = "cctu"))
dlu_b <- read.csv(system.file("extdata", "test_B_DLU.csv", package = "cctu"))
clu_a <- read.csv(system.file("extdata", "test_A_CLU.csv", package = "cctu"))
clu_b <- read.csv(system.file("extdata", "test_B_CLU.csv", package = "cctu"))

# Merge dataset with merge_data function
res <- merge_data(
  datalist = list(dt_a, dt_b), dlulist = list(dlu_a, dlu_b),
  clulist = list(clu_a, clu_b)
)
dt <- res$data # Extract combined data
dlu <- res$dlu # Extract combined DLU data
clu <- res$clu # Extract combined CLU data
```

## Apply macro dictionary

Next, we will apply the DLU and CLU files to the dataset. **Do not clean variable name** in the dataset and DLU/CLU files before applying `apply_macro_dict`. The function can handle variable name cleaning by setting `clean_names = TRUE`, this is default behaviour. The `apply_macro_dict` function will clean the variable name in the DLU and CLU, then apply these macro meta data to the target dataset. The cleaned DLU data will be stored internally and is further used by `cttab` to report missingness. You can use `get_dlu` to extract the internal DLU data after applying `apply_macro_dict` to the dataset for further usage or as a reference. Although you can use `set_dlu` to change the internal DLU data stored by `apply_macro_dict`, this is not recommended unless you know what you are doing. This will have an impact on the missing report.

```{r echo=TRUE}
# Create subjid
dt$subjid <- substr(dt$USUBJID, 8, 11)

# Apply CLU and DLU files
dt <- apply_macro_dict(dt, dlu = dlu, clu = clu, clean_names = FALSE)

# Give new variable a label
var_lab(dt$subjid) <- "Subject ID"
```

After this, you should follow the `Analysis Template` vignette and setup the population etc. 

```{r echo=FALSE}
# You should read
set_meta_table(cctu::meta_table_example)
# Create the population table
popn <- dt[, "subjid", drop = FALSE]
popn$safety <- TRUE

create_popn_envir("dt", popn)
.reserved <- ls()
```

Next we will do some data analysis.

## Data analysis

After you have attached the population, next thing you may want to do is extract a particular form from the data. For the table, assume we are reporting age, sex and BMI from the patient registration form by treatment arm. For demonstrating the filtering, we only report non-white patients' BMI. 

```{r echo=TRUE}
# Attach population
attach_pop("1.1")

# Extract patient patient registration form and keep subjid variable
df <- extract_form(dt, "PatientReg", vars_keep = c("subjid"))

# Now report Age, Sex and BMI. For BMI, report not white only
X <- cttab(
  x = c("AGE", "SEX", "BMIBL"), # Variable to report
  group = "ARM", # Group variable
  data = df, # Data
  select = c("BMIBL" = "RACEN != 1")
) # Filter for variable BMI

# Write table
X
```

## Formula interface

One can use formula in the `cttab` just like `lm`, but you will not be able to group variables (described in the later section). The left hand side of the formula is the variables to be summarised. Right hand side of the formula is the grouping and/or row splitting variables. The by visit variable should be separated by `|` with grouping variable, use `1` if there is no grouping variables. No `group` or `row_split` parameters to be used in the formula interface. All the other parameters are the same.
```{r eval=FALSE}
# No grouping
X <- cttab(AGE + SEX + BMIBL ~ 1, data = df)
# Group summarise by ARM
X <- cttab(AGE + SEX + BMIBL ~ ARM, data = df)
# Group summarise by ARM and split by visit cycle
X <- cttab(AST + BILI ~ ARM | AVISIT, data = df)
# Split by visit cycle
X <- cttab(AST + BILI ~ 1 | AVISIT, data = df)
```

## Missing data report

The `cttab` function will report the missing internally. You can use the following to get the missing report.

```{r echo=TRUE, eval=FALSE}
# This will save the missing report under Output folder
# Or you can set the output folder and name
dump_missing_report()

# Pull out the missing report if you want
miss_rep <- get_missing_report()

# Reset missing report
reset_missing_report()
```

After this, you can finish the remaining as in the `Analysis Template` vignette.

# More to cttab

As you have seen previously, the `cttab` function can easily populate simple tables. 

## Simple table

Table only some variables, no treatment arm or variable selection.
```{r echo=TRUE}
X <- cttab(
  x = c("AGE", "SEX", "BMIBL"), # Variable to report
  data = df
) # Data

X
```

## By group and filter

This is what we have seen before
```{r echo=TRUE}
X <- cttab(
  x = c("AGE", "SEX", "BMIBL"), # Variable to report
  group = "ARM", # Group variable
  data = df, # Data
  select = c("BMIBL" = "RACEN != 1")
) # Filter for variable BMI

X
```

## Split table row by visit

You can define `row_split` parameter to the name of visit or repeat variable.
```{r echo=TRUE}
attach_pop("1.1")
df <- extract_form(dt, "Lab", vars_keep = c("subjid", "ARM"))

X <- cttab(
  x = c("AST", "BILI", "ALT"),
  group = "ARM",
  data = df,
  row_split = "AVISIT", # Visit variable
  select = c("ALT" = "PERF == 1")
)

X
```

When `row_split` is supplied you can also flip the row hierarchy with the `nest` argument. The default `nest = "split"` puts the row-split levels on the outside (a "Study Visit = Baseline" banner with each variable's stats nested inside it). Setting `nest = "var"` flips it so each variable becomes the outer banner and the row-split levels (e.g. each visit) appear as sub-sections within it — useful when you want to read one variable across all visits at a glance.

```{r eval=FALSE}
X_var <- cttab(
  x = c("AST", "BILI", "ALT"),
  group = "ARM",
  data = df,
  row_split = "AVISIT",
  nest = "var"
)
```

## Group variable

In this example, we will report demographic variable, lab results and lab abnormality. Variables will be grouped, no group name will be given to demographic variables, "Blood" to lab results and "Pts with Abnormal" to lab abnormality. Here, we count the number of patients with abnormal lab results. The `cttab` will report the count and percentage of `TRUE`. This is useful if you want to report patient numbers for different condition that belong to one category. Below is how to do it:

```{r echo=TRUE}
# Prepare data as before
attach_pop("1.1")
df <- extract_form(dt, "PatientReg", vars_keep = c("subjid"))
base_lab <- extract_form(dt, "Lab",
  visit = "SCREENING",
  vars_keep = c("subjid")
)

# Define abnormal
base_lab$ABNORMALT <- base_lab$ALT > 22.5
var_lab(base_lab$ABNORMALT) <- "ALT abnormal"
base_lab$ABNORMAST <- base_lab$AST > 25.5
var_lab(base_lab$ABNORMAST) <- "AST abnormal"

df <- merge(df, base_lab, by = "subjid")

# Table
X <- cttab(
  x = list(c("AGE", "SEX", "BMIBL"),
    # Group lab variable
    "Blood" = c("ALT", "AST"),
    # Group abnormal variable
    "Pts with Abnormal" = c("ABNORMAST", "ABNORMALT")
  ),
  group = "ARM",
  data = df,
  # Add some filtering
  select = c(
    "BMIBL" = "RACEN != 1",
    "ALT" = "PERF == 1"
  )
)

X
```

## Rounding

The default behaviour of this function is to keep one digits for the percentage and 3 significant value for the numerical values. The default rounding function is `signif_pad`, you can also use `round` or `round_pad` to keep digits in the summary. The `format_percent` function is used to format the percentage values. There is `format_pval` might be useful to you.

```{r echo=TRUE}
X <- cttab(
  x = c("AGE", "SEX", "BMIBL"), # Variable to report
  group = "ARM", # Group variable
  data = df, # Data
  digits = 2, # Keep 2 digits for numerical
  digits_pct = 1, # Keep 1 digits for percentage
  rounding_fn = round
) # Use function round for rounding

X
```

## What `cttab()` returns and how it renders

The print method you've been using above does a small amount of work behind the scenes. `cttab()` itself returns a *long-format* `data.table` (one row per `(row_split, group level, variable, statistic)` cell) carrying the raw values, before any layout decisions are made:

```{r echo=TRUE}
X_long <- cttab(c("AGE", "SEX"), group = "ARM", data = df)
class(X_long)
head(X_long, 6)
```

This is the shape you want when you need to post-process the numbers — filter rows, edit labels, or bind two tables together (`rbind.cttab()` is aware of the long format and re-numbers `Var_ID` / `Group_ID` so the bound parts stay in their own sections). It is also what `write_table()` accepts when given an unformatted `cttab`.

The rendered shape — section banners inserted, headers stamped with bold/bgcol/span tokens, stat rows indented — is produced by `cttab_format()`. `print.cttab()` and `write_table()` call it for you, so you normally don't have to. Calling it explicitly is useful when you want to inspect or tweak the rendered layout:

```{r echo=TRUE}
X_fmt <- cttab_format(X_long)
class(X_fmt)                   # c("cttab", "data.frame")
head(X_fmt[, "label", drop = FALSE], 8)
attr(X_fmt, "row_style")[1:8]
```

The result is a `data.frame` with a `label` column carrying the per-row text and a `row_style` attribute carrying the per-row `;`-joined style tokens — the same vocabulary `format_table()` exposes (see the next section). A few notes:

* `cttab_format()` honours the `nest` attribute on `X_long`, so flipping `nest = "split" ↔ "var"` is reflected in the rendered row order without a second function call.
* If `cttab_format()` is handed a table that has already been formatted (a matrix `cttab` or any data.frame stamped with a `row_style` attribute and a `label` column), it returns the input unchanged. That makes it safe for `write_table()` and `print.cttab()` to call unconditionally.
* If you only want a tidy long-format frame for downstream tools (no rendering metadata), use `as.data.frame(X_long)` — it drops the `Group_ID`, `Group_Label`, `Var_ID`, `Stat_ID` book-keeping columns and keeps `Variable`, `Statistic`, `Value`, plus the `group` / `row_split` columns.

## Styling a custom table with `format_table`

Sometimes the table you want to render isn't a `cttab` summary at all — maybe you've built a small `data.frame` by hand, or run a regression and want to drop a tidy version into the report. `write_table()` knows how to render any `data.frame` or `matrix` styled with a `row_style` attribute, and `format_table()` is the helper that stamps that attribute on for you.

The styling vocabulary is four tokens — `bold`, `bgcol`, `span`, `indent` — plus an optional `col` token for coloured text. You pass each one a vector of row indices, and `format_table()` assembles the per-row `;`-joined string that the renderer consumes. This is the same machinery `cttab_format()` uses internally, so the look will match a `cttab` table.

```{r echo=TRUE}
df <- data.frame(
  label = c("Demographics",
            "Age (years)", "Mean (SD)", "Median [Min, Max]",
            "Sex",         "Female",    "Male"),
  Treated = c("",          "",          "65.2 (8.1)", "66 [50, 82]",
                           "",          "12 (40%)",   "18 (60%)"),
  Control = c("",          "",          "63.9 (7.4)", "64 [49, 80]",
                           "",          "15 (50%)",   "15 (50%)"),
  stringsAsFactors = FALSE
)

df <- format_table(df,
  bold   = c(1, 2, 5),  # banner + variable headers
  bgcol  = 1,           # only the top "Demographics" banner gets a fill
  span   = c(1, 2, 5),  # banner + variable headers span all columns
  indent = c(3, 4, 6, 7)
)

attr(df, "row_style")
```

A row that should appear as a section banner ends up with `"bold;bgcol;span"`; a plain variable header is `"bold;span"`; stat rows under it are `"indent"`; everything else stays `""`. Token order in the assembled string is fixed (`bold;bgcol;span;indent;col`), and out-of-range / `NA` row indices are rejected with the column name in the error message.

For coloured fills or text, pass `bgcol_color` / `col_color`. Both accept either a six-digit hex string or any standard R colour name; named colours are converted to hex once so the XSLT pipeline only sees one form:

```{r echo=TRUE, eval=FALSE}
df <- format_table(df,
  bgcol       = 1,
  bgcol_color = "steelblue", # banner uses a custom fill
  col         = c(3, 4),
  col_color   = "#B22222"    # two stat rows in a deep red
)
```

`col` requires `col_color` to be set — supplying one without the other emits a warning and the colour token is dropped. If you want the package default (a light grey fill), just omit `bgcol_color`.

Once stamped, hand the table off to `write_table()` (or the print method) the same way you would with a `cttab`:

```{r eval=FALSE}
write_table(df)   # renders into Output/Core/table_<number>.xml
```



